I have many people to thank for helping me get through the past four years in
overall pretty good shape. First I have to thank my main advisor, Nadav Bar,
for hiring me and thereby giving me a chance to see what the world of research
is all about.

Second of all I would like to thank my fellow PhD students at NTNU for their
help and friendship. From Chemical Engineering I thank Deeptanshu for all the
good discussions and culinary experiences, Johannes for talking about window
managers and climbing, Bjørn-Tore for many, many climbing sessions, Ivan for
teaching me \textit{rude}mentary Italian while finishing his thesis, Olaf for
insisting that I start using Linux, and Magnus, Maryam and Ismael for good
company and discussions. From Biotechnology I thank Rahmi for suggesting I
should check out Python, Veronika for good collaboration and companionship, and
Simone and Friederike for their collaboration. From Biology I thank Aravind and
Konika for many conversations and for sharing very good Indian food.

Thirdly I owe a great \textit{moltes gràcies} to Roderic Guigó and his group at
the CRG in Barcelona. I came to you during a challenging time of my PhD and you
took me along from the very beginning. Staying at the CRG was an eye-opener for
me in terms of what what a high-level, productive and professional research
environment looks like. Thanks to Roderic, Pedro, Rory, Sarah, Angelika, David,
Maik, and all the rest for teaching me about bioinformatics and genomics.

Fourthly I thank Martin Kuiper for becoming my co-supervisor during the last
year of my PhD and for helping me getting through to the end. Your help was
essential when I made the transition from the CRG back to the NTNU and also in
the final stages of writing the thesis.

Fifthly, I want to give a very large \textbf{thank you} to Lilian M. Hsu.
Through so many emails Lilian has patiently and constructively helped me
understand about the intricacies of transcription initiation. I thank you for
your help and your collaboration. I especially want to thank you for trusting
my results enough to perform experiments using the ITS-sequences I had
suggested. That was one of the definite highlights of my PhD. Your help has
lifted my understanding of transcription initiation, and thereby this thesis,
considerably.

I am also happy to thank my mother and my father in Tromsø and my sister now in
Oslo. You have always supported me and made my life easy so that I could focus
on the things I wanted to do. Thank you so much for that.

Finally and most importantly I thank my wonderful Itziar for her ever-positive
spirits, happy smiles, and for reminding me time and time again about what is
important in life. Thank you for listening to my ramblings about some topic or
another in molecular or evolutionary biology and for patiently explaining all
the things I never learned. Thank you for making life easy and happy and for
inspiring me with your imagination and creativity.
