% This is just for letting me cite stuff.
%\addbibresource{/home/jorgsk/phdproject/bibtex/jorgsk.bib}
\subsubsection{Overview of transcription in eukaryotes}
In this section we will review the process of transcription termination in
eukaryotes, and more specifically in mammalian cells. We will in general speak
about eukaryotic transcription, but when something stands out in mammals
compared to other eukaryotes, such as yeast, we will explicitly say that we are
talking about results obtained from mammalian cells. Since the previous
sections were about transcription and translation in bacteria, we will often
make the contrast between transcription in eukaryotes and in bacteria.

In eukaryotes there are several types of RNA polymerases, each one responsible
for transcribing different classes of genes. The most abundant polymerase is
RNAP II which transcribes mostly protein coding genes. With 12 subunits in
mammals, RNAP II is larger than its bacterial counterpart, but the core
subunits are structurally and functionally conserved compared to bacterial RNAP
\cite{ebright_rna_2000}.

As in bacteria, transcription in eukaryotes begins when RNAP II is recruited by
transcription factors to promoter sites, although this recruitment is more
diversified than in bacteria since no equivalent to the sigma factor system
exists in eukaryotes. Transcription initiation also involves the same steps of
open bubble formating and abortive cycling, although additional assisting
transcription factors are involved compared to bacteria
\cite{wade_transition_2008}. Transcription elongation by RNAP II also involves
pausing and backtracking, but pausing on eukaryotic DNA is more complicated due
to wrapping of DNA in nucleosomes and chromatin \cite{sims_elongation_2004}.

Transcription termination on the other hand is markedly different between
eukaryotes and bacteria. In bacteria, the position where RNAP stops
transcribing also marks the genomic position of the 3\p end of the mRNA. In
eukaryotes however, transcription termination does not co-occur with mRNA
release and 3\p end definition. Instead, mRNA is released by a process of
cleavage and polyadenylation several hundred nucleotides before RNAP terminates
transcription.

\subsubsection{Cleavage and polyadenylation marks the 3' end of mRNA
and promotes transcription termination}
There are usually several sites where cleavage and polyadenylation can occur in
a gene, though these sites are predominantly found in the 3\p UTR. The cleavage
and polyadenylation sites are identified by the polyadenylation signal (PAS),
which is a highly conserved AATAAA signal sequence or a close variant like
ATTAAA. At some sites, there is also an enhancing U or GU rich sequence further
downstream the PAS. The CPSF (cleavage/polyadenylation specificity factor)
protein binds to the PAS along with several other factors to form a poly(A)
binding complex. The complex then cleaves mRNA at a position between 10 and 30
nucleotides downstream the PAS (see \cite{proudfoot_ending_2011} for a
comprehensive and historical review of 3\p cleavage and polyadenylation). As
soon as mRNA has been cleaved, a poly(A)-polymerase will add a polyA tail
(length 250 in mammals) to the 3\p end and the tail will be covered by poly(A)
binding protein (PAB). The PAB protects the poly(A) tail from degradation and
assists in transporting the mRNA to the cytoplasm
\cite{mangus_poly-binding_2003}.

As cleavage and polyadenylation occur, RNAP II is still transcribing the
downstream DNA. It is important that RNAP II terminates transcription,
otherwise it can collide with the polymerases that transcribe downstream genes.
One characterized method of transcription termination is by a 5\p to 3\p
exonuclease called XRN2. In what is called the ``torpedo model'', this
exonuclease degrades the still-transcribed RNA from the 5\p end until it
reaches RNAP II and causes transcription termination, although the exact
details of how transcription termination happens is not clear
\cite{kuehner_unravelling_2011}.

\subsubsection{mRNA processing is necessary step before translation}
In contrast to bacterial transcripts, eukaryotic mRNA needs extensive
processing before it is capable of being translated by ribosomes. Therefore,
mRNA in eukaryotes is called pre-mRNA until it has gone through the necessary
processing steps. There are essentially three main events that make up pre-mRNA
processing: 5\p cap addition, splicing, and 3\p cleavage and polyadenylation.

The 5\p cap is a guanine nucleotide variant that is added to the 5\p nucleotide
of a transcript shortly after it emerges from the RNA exit channel of RNAP II.
The 5\p cap presumably protects against 5\p exonucleases and is necessary for
the proper export of the mRNA from the nucleus.

Splicing is the removal of introns, the non-coding parts of pre-mRNA that make
up most of gene sequences in mammals. By splicing out the introns and joining
the surrounding exons, only the coding regions (as well as the 3\p and 5\p
untranslated regions) of pre-mRNA make up the final mRNA.

Several proteins have shared roles across the different pre-mRNA processing
steps. In other words, there is cross-talk between the processing pathways. For
example, the presence of the 5\p cap increases the efficiency of the excision
of the 5\p proximal intron and the efficiency of polyadenylation; and
polyadenylation again increases the efficiency of the excision of the 3\p
terminal intron \cite{proudfoot_integrating_2002}. As a concrete example of
cross-talk, it was recently found that the U1 ribonucleoprotein, which was
previously known for its role in splicing, prevents premature cleavage and
polyadenylation at cryptic polyadenylation sites \cite{kaida_u1_2010}.

If any of the processing steps are inefficiently executed or not executed at
all, the pre-mRNA transcript will be targeted for degradation either in the
nucleus itself or in the cytoplasm after transport \cite{doma_rna_2007}. This
reduces the risk that errors during transcription and pre-mRNA processing will
result in nonfunctional or possibly harmful protein products.

\subsubsection{Usage of alternative polyadenylation sites}
% 1) For making different 3 UTR
Each 3\p UTR region of a gene often has several alternative sites where
cleavage and polyadenylation may happen \cite{tian_large-scale_2005}. Depending
on which of these sites are used for cleavage and polyadenylation, the length
of the 3\p UTR in the final mRNA will vary. This can affect the downstream fate
of the mRNA as the 3\p UTR region contains many regulatory elements, such as
binding sites for microRNA. MicroRNAs that bind the 3\p UTR region in the
cytoplasm generally decrease expression from the mRNA they bind to. It was
previously unclear whether the microRNAs decreased expression by inducing
transcript degradation or simply by blocking translation. Recently, it was
established that at least in mammals microRNA in the 3\p UTR decrease
expression by causing the transcript to be degraded
\cite{huntzinger_gene_2011}. Since microRNAs are involved in a host of
metabolic processes and human diseases \cite{huang_biological_2010}, the choice
of where to cleave and polyadenylate a transcript can have wide-spanning
consequences.

There are also examples of cleavage and polyadenylation in sites outside the
3\p UTR. The most prominent of these are sites within introns. The use of an
intronic cleavage and polyadenylation site a site would cause any downstream
exons to be left out of the transcript, resulting in a shorter peptide sequence
upon translation. Thus, the site of cleavage and polyadenylation can also
modulate the protein coding content of the mRNA. A well known example is from
the immunoglobin protein in B cells. Depending on whether a polyadenylation
site in an intron is used or not, the membrane bound or the secreted version of
this protein is made \cite{peterson_regulated_1989}.

\subsubsection{Transient polyadenylation unrelated to mRNA processing}
In the last decade, it has become clear that there is polyadenylation of
RNA in eukaryotic cells that is unrelated to mRNA 3\p processing. First, it was
discovered that poly(A) tails were added to aberrant transcripts in the nucleus
of yeast. The protein complex responsible for this polyadenylation was given
the name TRAMP, and transcripts polyadenylated in this way were found to be
targeted for degradation by the nuclear exosome \cite{lacava_rna_2005,
wyers_cryptic_2005}. This was a surprising and important discovery, as
previously degradation-related polyadenylation was only known from bacteria,
prompting the suggestion that degradation-associated polyadenylation by TRAMP
has been conserved in the nucleus from the original bacterial degradation
machinery \cite{lacava_rna_2005}. Later, degradation-related polyadenylation
was found in the nucleus of mammals too, and eventually even in the cytoplasm
of human cells \cite{slomovic_polyadenylation_2006, slomovic_addition_2010}. In
summary, there is an emergent role for, degradation-related
polyadenylation of some RNA species in eukaryotes that was previously only
thought to occur in bacteria. In contrast, the poly(A) tail formed on pre-mRNA
protects the mRNA against degradation. One possibility why the new form of
poly(A) tail does not confer stability as well is that this tail is too short
for the PAB to bind \cite{lacava_rna_2005}, although the exact length of these
poly(A) tails was not determined.

\subsubsection{Genome-wide studies of polyadenylation}
The study of sites of polyadenylation across the genome has occurred in three
stages in the last 12 years, with each stage resting on a different type of
technology. The first wave used cDNA and EST sequence data obtained by
laborious Sanger sequencing. The second wave used microarray and SAGE
technologies, and the third wave used the RNA-seq next generation sequencing
technology. We will review key results obtained in these three stages
chronologically.

From the early 90s onward, more and more human expressed sequence tags (ESTs)
from cDNA became available. The increasing amount of sequence data facilitated
for the first time large scale analysis of 3\p UTRs and polyadenylation sites.
The polyadenylation site of an EST is found by identifying a poly(A) or poly(T)
extremity which does not correspond to a genomic sequence, trimming that
extremity, and matching the remainder to template mRNA or genomic sequence
\cite{beaudoing_patterns_2000, tian_large-scale_2005}.

These early genome-wide studies were successful in determining i) the frequency
of occurrence of the different PAS \cite{beaudoing_patterns_2000}, ii) that
over half of human genes employ alternative polyadenylation, iii) that sites of
alternative polyadenylation are evolutionary conserved between humans and mice
\cite{tian_large-scale_2005}, and vi) that there is extensive polyadenylation
at intronic polyadenylation sites \cite{tian_widespread_2007}.

However, EST data limits the type of questions that can be investigated. First,
EST data was in low quantity, due to the expenses and time of Sanger
sequencing. Thus, the only way to compare alternative polyadenylation on a
genome-wide scale was to include EST data from different experiments from
different cell lines and tissues. This had the consequence that our literature
review revealed no studies with \textit{de novo} EST sequencing for the purpose
of studying polyadenylation. All studies used EST sequences from databases.
Further, the EST data is biased toward protein coding genes that were found
interesting enough to sequence individually. Thus many classes of
polyadenylated RNA, such as long noncoding RNA, were possibly missed by these
studies. Finally, although EST data may be used to give a quantitative profile
of gene expression, the output data is often normalized so that the
quantitative profile is lost. It therefore difficult to compare expression
values across genes with EST data, although some approaches have been developed
for this purpose \cite{liu_quantitative_2006}.

As the microarray technology matured and more full-length genomes became
available, microarrays were used to study 3\p UTR variation and
polyadenylation, often in combination with EST data. Due to the low price and
high speed of microarrays, it was possible to use two microarray experiments in
the same study to compare the usage of alternative 3\p ends. In practise, to
test for differential 3\p UTR usage, microarray probes corresponding to the
normal or extended 3\p UTR are used, and signal intensities from the probes
under different conditions can be compared to see if a 3\p UTR is longer or
shorter under certain conditions \cite{sandberg_proliferating_2008,
ji_progressive_2009}. With microarrays, 3\p UTR expression levels could be
compared across these experiments. This allowed for a broader range of
hypotheses to be tested. The lengthening or shortening of a 3\p UTR would
correspond to the choice of upstream or downstream polyadenylation sites.

Results obtained with microarray and EST data include different patterns of
polyadenylation in different human tissues \cite{zhang_biased_2005}, and
wide-spread shortening of 3\p UTR length during immune cell activation
\cite{sandberg_proliferating_2008}. A combination of EST, microarray, and SAGE
data showed a progressive lengthening of mouse 3\p UTRs during embryonic
development \cite{ji_progressive_2009}.

Microarray studies, although capable of providing a wealth of insight, are
fundamentally limited in scope when the purpose is to study alternative
polyadenylation. This is because no direct evidence of polyadenylation sites
can be obtained with microarrays, since the information is not in
sequence-format. With microarrays, one can only investigate those annotated
genes which are already suspected or known to be subject to alternative
polyadenylation. Further, microarray output is only in terms of relative
differences of gene expression. This means that it is not possible to compare
3\p UTR lengths across different genes.

With the advance of second generation sequencing technologies in the late 2000s
in the form of RNA sequencing (RNA-seq), many of the limitations of both EST
and microarray data seem to have been resolved. RNA-seq combines the best of
EST and microarray data when studying alternative polyadenylation. Firstly,
like ESTs, the RNA-seq data is in sequence format, allowing the direct
detection of poly(A) tails and thereby the site of cleavage and
polyadenylation. Secondly, like EST data, RNA-seq data is quantitative,
allowing the direct comparison of expression levels of 3\p UTRs across the
genome. And thirdly, like microarrays, RNA-seq can easily be performed on RNA
samples from under different conditions in the same experiment, allowing direct
hypothesis testing which is impossible when using only information from
databases. 

RNA-seq was rapidly used to study the polyadenylation landscape for cell lines
and tissues. These experiments confirmed what had been discovered earlier by
single-mRNA studies and EST analysis; that AAUAAA is the canonical
polyadenylation signal, that single genes can be represented with multiple
sites of polyadenylation, and that there is frequent polyadenylation of
intronic sequences. The new discoveries included first of all many novel
polyadenylation sites scattered across the genome
\cite{ozsolak_comprehensive_2010, derti_quantitative_2012}. It was also found
that intronic and intergenic polyadenylation sites are in humans associated
with a novel TTTTTTTTT motif which does not occur at the normal polyadenylation
sites in 3\p UTRs \cite{ozsolak_comprehensive_2010}. Further, genome wide
annotation of polyadenylation sites from sense and antisense transcription was
obtained in large scale for the first time in \textit{C. elegans} and
\textit{Arabidopsis} \cite{mangone_landscape_2010, wu_genome-wide_2011}. The
last examples show how RNA-seq is directly used in genome annotation.

As mentioned, microarray studies revealed unidirectional changes of 3\p UTR
length during cell development stages \cite{sandberg_proliferating_2008,
ji_progressive_2009}, hinting that 3\p UTR lengths can be regulated at a global
level. A study that did not use high-throughput data identified shortening of
the 3\p UTR of many transcripts in a cancer cell line, proposing that this was
a general characteristic in cancer cells \cite{mayr_widespread_2009-2}. As a
follow up to this study, Fu et al. compared the relative change in 3\p length
between two cancer cells and an epithelial cell line
\cite{fu_differential_2011}. They did not find a consistent pattern of
shortening of 3\p UTRs in the cancer cell lines. Instead, the 3\p UTR lengths
of one of the cancer cells was shorter and 3\p UTR lengths of the other was
longer than the epithelial cell. This suggests that there is no clear-cut
genome-wide trend of short 3\p UTRs in cancer cells, contrary to what had
previously been concluded.

An unexpected finding with RNA-seq has been the detection of poly(A) tails for
histone mRNAs in both human, mice, and \textit{C. elegans}
\cite{mangone_landscape_2010-1, shepard_complex_2011}.  Histone mRNAs were
previously thought to be the only mRNAs in metazoans without a poly(A) tail
\cite{marzluff_metabolism_2008}, even though several of the histone genes had
been found to contain the AATAAA polyadenylation signal at the 3\p end
\cite{keall_histone_2007}. A possible explanation for the discovery of histone
mRNA with poly(A) tails is that the histone transcripts are first cleaved and
polyadenylated, and subsequently processed to lose their poly(A) tail
\cite{mangone_landscape_2010-1} (See supplementary materials). Another option
is that if some histone mRNA are polyadenylated, they avoid degradation at the
end of the S-phase, and may thus be translated throughout the cell cycle
\cite{shepard_complex_2011}.

The most recent and thorough study of genome-wide polyadenylation was done
using a novel sample-preparation protocol by Derti et al. They used RNA-seq to
find polyadenylation sites in five mammal species, including human, in 24
tissues \cite{derti_quantitative_2012}. They found over 400.000 polyadenylation
sits in the human tissues, compared to 150.000 found previously. However, how
many of these sites correspond to normal mRNA 3\p stable poly(A) tail or
degradation-related transient poly(A) tails was not investigated. On reason for
this very large number of sites could be that they identified most novel
polyadenylation sites in lowly expressed transcripts, which may not previously
have been detected. Derti et al also found that although many poly(A) sites
were tissue specific, 70 \% of genes showed the same usage of alternative
polyadenylation across all tissues.

\subsubsection{Preparing a cDNA library for RNA-sequencing}
The work presented in this thesis that deals with polyadenylation is based on
sequencing data made with Illumina sequencing technologies. Here we briefly
mention the stages of an RNA-seq experiment that are relevant for interpreting
the subsequent results.

To obtain good sequencing results, fragmented DNA in large concentrations is
required. To reach these concentration, it is generally necessary to both
amplify the cellular RNA extract and to convert the RNA to cDNA by reverse
transcription. Both RNA amplification and cDNA conversion can introduce biases
depending on the choices of methods and the order in which they are performed
\cite{wang_rna-seq:_2009}. From the point of view of studying polyadenylation,
it is necessary that the poly(A) tail is preserved during cDNA library
preparation. This requires the use of poly(T) primers both for making the cDNA
library and for PCR amplification.

After cDNA preparation and several rounds of PCR amplification, most poly(A)
tails will be reduced in length from the full 250 nt. This is because the
poly(T) primers will bind randomly in the poly(A) tail, something which will
gradually reduce the average length of the poly(A) tails.

\subsubsection{Sequencing errors and biases}
There are several errors and biases that creep into sequencing that must be
taken into account during the computational analysis of the output. Here we
discuss one error and one bias that is important for analysing Illumina short
pair ended reads.

An error that is particularly important for the study of polyadenylation is
that sequencing performance degrades over single-nucleotide stretches
\cite{minoche_evaluation_2011}. This implies that when sequencing the poly(A)
sequence, other nucleotides than just A will be reported with a relatively high
frequency.. This means that when looking for poly(A) stretches in the output as
signatures of poly(A) sites, some variation from the homopolymer sequence must
be expected.

An important bias is that most sequences with poly(A) tails will be output in
the reverse-transcribed version. That is, 5\p CCCGAAAA 3\p will be output as 5\
TTTTCGGG 3\p. This is due to the RNA fragmenting and that sequencing
(performed by a DNA polyermase) happens in the 5\p to 3\p direction and that
each RNA is present in both its original and reverse transcribed form. During
sequencing, 50 to 150 nucleotides will be sequenced from each side of the around
300 nt fragment. If the poly(A) tail is the 50 last nucleotides of the 300 nt
fragment RNA, sequencing 150 nucleotides will never reach nucleotide 250 where
the poly(A) site is. However, the reverse transcribed version contains 50 Ts at
the 5\p end, which ensures that sequencing this version of the RNA gets the
whole poly(T) tail as well as 100 bases which can be mapped to the genome.

Even though the actual output poly(A) tail is in the reverse transcribed
poly(T) format, we will often refer to both of them as a poly(A) tail.

\subsubsection{Mapping reads to the genome}
The sequence-snippets that are output from sequencing machines are called
reads, and generally come in sizes from 30 to 500 basepairs, depending on the
technology used. If a reference genome exists for the organism from which the
RNA sample was taken, these reads must be mapped to that genome. Otherwise, it
is possible to attempt to construct the organisms transcriptome \textit{de
novo} from the RNA-seq output, although this generally gives poor results.

Mapping read to the genome means trying to find where the RNA-snippet from the
sequencing machine originated from. Different methods are used for this, but
what they have in common is that they allow for mismatches between the read and
the genome to allow for sequencing errors. Extra care is taken to map reads
spanning the intron-exon junctions, since these reads must necessarily be split
in two to be mapped, possibly thousands of nucleotides apart.
