I have many people to thank for helping me get through the past years in good
shape. First I owe a great thanks to my main advisor, Nadav Bar, for taking me
on as a student and giving me a chance to see what the world of research is all
about.

Second of all I would like to thank my fellow PhD students at NTNU for their
help and friendship. From the Department of Chemical Engineering I thank
Deeptanshu for all the meaningful discussions and culinary experiences,
Johannes for talking about window managers and climbing, Bjørn-Tore for many,
many climbing sessions, Ivan for teaching me \textit{rude}mentary Italian while
writing up his thesis, Olaf for insisting that I start using Linux from day
one, and Magnus, Maryam and Ismael for good company and discussions. From
Biotechnology I thank Rahmi for good advice and suggesting I should check out
Python; I thank Veronika for friendship and good collaboration, and Simone and
Friederike for many interesting discussions. From the Department of Biology I
thank Aravind and Konika for many conversations and for sharing very good
Indian food during lunch. An extra thanks to Konika for making it so that I
could stay with her family in Delhi when I was there -- it was a blast!
Finally, I thank Deeptanshu, Aravind, and Shalini for the many shared dinners,
cabin trips and good conversations. You made this journey easier!

Thirdly I owe a great \textit{moltes gràcies} to Roderic Guigó and his group at
the CRG in Barcelona. I came to you during a challenging time in my PhD work
and you took me along from the very beginning. My stay in Barcelona was an
eye-opener, as I observed and partook in how a high-level, productive and
professional research environment like the CRG operates. Thanks to Roderic,
Pedro, Rory, Sarah, Angelika, David, Maik, and all the rest for teaching me
about bioinformatics and genomics.

Fourthly I thank Martin Kuiper for becoming my co-supervisor during the last
years of my PhD and for helping me getting through to the end. Your help was
especially appreciated when I moved from the CRG back to NTNU and also in the
final stages of writing the thesis.

Fifthly, I want to give a very large \textit{thank you} to Lilian M. Hsu.
Through so many emails Lilian has constructively aided my education of the
intricacies of transcription initiation. I thank you for your help and your
collaboration. I especially want to thank you for trusting my modeling results
enough to perform experiments using the ITS-sequences I had suggested. Waiting
for the outcome those experiments was definitely the most exciting part of my
PhD.

I am also happy to thank my mother and my father in Tromsø and my sister in
Oslo. You have always supported me and made my life easy so that I could focus
on the things I wanted to do. Thank you so much for that.

Finally and most importantly I thank my wonderful Itziar for her positive
spirits, happy smiles, and for reminding me time and time again about what is
important in life. Thank you for listening to my ramblings about my latest
fancy in molecular or evolutionary biology and for patiently explaining all the
things I never learned. Thank you for making life easy and happy and for
inspiring me with your imagination and creativity!
