%\addbibresource{/home/jorgsk/phdproject/bibtex/jorgsk.bib}

\section{Utail pipeline overview}

Utail takes the mapped RNA-seq data through the following pipeline:
\begin{enumerate}
		%1
	\item (Only if annotation was supplied) Create a list of all possible
		3'UTR regions and polyadenylation sites from the genome annotation
		%2
	\item Filter the RNA-seq reads and retain only the unmappable reads
		%3
	\item Further filter the unmappable reads to keep only those with leading
		poly(T) or trailing poly(A) stretches
		%4
	\item Remove the poly(T) or poly(A) stretch from the read and remap the
		remaining part to the genome -- keep those that map uniquely
		%5
	\item Determine the site of polyadenylation depending on poly(A/T)
		including the strand
		%6
	\item Cluster together all reads that map within 40 nt of each other
		%7
	\item For each cluster, get the genomic sequence up to 60 nt downstream
		from the polyadenylation site. Note any PAS and discard cluster if a
		genomic poly(A) region is found at the site of the cleaved poly(A)
		stretch.
		%8
	\item For each poly(A) cluster, write to the output
\end{enumerate}
We will now elaborate on each the above points.\\

\textbf{1.)} By providing an annotation of the transcripts from the genome
re-map the read back to the genome. If the read maps uniquely (with up to 2
mismatches), consider this the 3' end of the mapping a putative poly(A) site.

\textbf{2.)} With default settings, short-read mappers will not be able to map
to the genome a read-fragment containing the non-genomic poly(A)
stretch. Therefore, we are only interested in those reads which were initially
unmapped. The output format of mappers usually contains a flag which says if
the read was mapped or not.

\textbf{4.)} Among the unmapped reads, we are only interested in the ones which
contain a stretch of poly(A) at the 3' end or poly(T) at the 5' end. A minimum
of 5 poly(A/T) or 5 poly(A/T) in 7 was used.

Once the poly(A) tail has been removed, we use the GEM mapper to re-map the
read back to the genome. If the read maps uniquely (with up to 2 mismatches),
consider this the 3' end of the mapping a putative poly(A) site.

\textbf{5.)} Even though the sequencing protocol is not strand-specific, we can
use the information from the poly(A) versus poly(T) stretch to find out which
genomic strand the read originated from. For example, if the read maps to the
+ strand with a poly(A) tail, the read came from the + strand. If the
read maps to the + strand with a poly(T) tail, the read originated from the -
strand. This allows the strand-specific determination of poly(A) sites.

\textbf{6.)} Since the exact site of cleavage and polyadenylation is stochastic,
we must cluster the different poly(A) sites. The average of this clustering is
reported as the cleavage and polyadenylation site for all reads in the cluster.

\textbf{7.)} The sequence around the putative poly(A) site is obtained for two
reasons. The first is to find PAS sequences downstream the site. The second is
to look for pol(A) or poly(T) stretches that could identify this site as a
false positive.

\textbf{8.)} The output for each cluster contains information about the
chromosome, strand, number of reads in the cluster, if the site has been
annotated, and if there is a PAS withing 40 nucleotides and the distance to
this PAS.
